---
title: "ST 558 Final Project: Exploratory Data Analysis"
author: "Lee Bennett"
format: html
editor: visual
---


```{r load packages}
#| include: false
library(tidyverse)
```

## Introduction

The purpose of this analysis is to create a predictive model for the incidence of diabetes using data from the 2015 [Behavioral Risk Factor Surveillance System (BRFSS)](https://www.cdc.gov/brfss/index.html) survey conducted by the Centers for Disease Control. The dataset we will use for modeling includes survey responses from 253,680 individuals.

The response variable in the dataset, `Diabetes_binary`, is a 0/1 indicator for the presence of diabetes or prediabetes. There are a number of potential predictors in the dataset, but we will consider only the following health-related variables in the dataset:

-   Binary factors for hypertension and high cholesterol (yes vs. no)
-   Reported body mass index (BMI)
-   Sex (male vs. female)
-   A factor for having any kind of health insurance coverage (yes vs. no)
-   An ordinal factor for self-rated general health, ranging from 1 ("Excellent") to 5 ("Poor")

## Analysis Data

We'll first read in the raw data from the CSV file and then transform the predictors of interest into a form suitable for analysis:

```{r create data}
#| warning: false

#Read in the raw data
diabetes_raw <- read_csv(file="diabetes_binary_health_indicators_BRFSS2015.csv",show_col_types=FALSE)par

#Create an analysis dataset containing only the relevant variables
yn_label <- c("No","Yes")
diabetes_ad <- diabetes_raw |> select(Diabetes_binary, HighBP ,HighChol, BMI, Smoker, Sex, AnyHealthcare, GenHlth) |>
  mutate(high_bp = factor(HighBP, levels=c(0,1), labels=yn_label),
         high_chol = factor(HighChol, levels=c(0,1), labels=yn_label),
         sex = factor(Sex, levels=c(0,1), labels=c("Female","Male")),
         smoker = factor(Smoker, levels=c(0,1), labels=yn_label),
         healthcare = factor(AnyHealthcare, levels=c(0,1), labels=yn_label),
         gen_health = factor(GenHlth, levels=c(1,2,3,4,5), labels=c("Excellent", "Very good", "Good", "Fair", "Poor")),
         .keep="unused")

#Check for missing data
colSums(is.na(diabetes_ad))
```
There is no missingness among the variables that we've chosen, so we can move on to explore the relationships between the incidence of diabetes and the predictor variables we've chosen.

## Exploratory Analysis

```{r eda 1}
#
diabetes_g1 <- diabetes_ad |> pivot_longer(!Diabetes_binary, names_to="var_name")
                                           
```



## Link to modeling page

[Click here for the modeling page.](Modeling.html "Modeling.html")
