---
title: "ST 558 Final Project: Exploratory Data Analysis"
author: "Lee Bennett"
format: html
editor: visual
---


```{r load packages}
library(tidyverse)
```

## Introduction

The purpose of this analysis is to create a predictive model for the incidence of diabetes using data from the 2015 [Behavioral Risk Factor Surveillance System (BRFSS)](https://www.cdc.gov/brfss/index.html) survey conducted by the Centers for Disease Control. The dataset we will use for modeling includes survey responses from 253,680 individuals.

The response variable in the dataset, `Diabetes_binary`, is a 0/1 indicator for the presence of diabetes or prediabetes. There are a number of potential predictors in the dataset, but we will consider only the following health-related variables in the dataset:

-   `HighBP`: a binary indicator for hypertension
-   `HighChol`: a binary indicator for hyperlipidemia
-   `HvyAlcoholConsump`: a binary indicator for alcohol consumption that meets or exceeds 14 drinks/week for men or 7 drinks/week for women
-   `AnyHealthcare`: a binary indicator for having any kind of health insurance coverage
-   `GenHlth`: an ordinal variable for self-rated general health, ranging from 1 ("Excellent") to 5 ("Poor")

Sex and age will also be considered.

## Analysis Data

We'll first read in the raw data from the CSV file and then transform the predictors of interest into a form suitable for analysis:

```{r create data}
#Read in the raw data
#| warning: false
diabetes_raw <- read_csv(file="diabetes_binary_health_indicators_BRFSS2015.csv")


#Create an analysis dataset containing only the relevant variables
diabetes_ad <- diabetes_raw |> select(Diabetes_binary, HighBP ,HighChol, BMI, Smoker, HvyAlcoholConsump, AnyHealthcare, GenHlth) |>
  mutate(high_bp = as_factor(HighBP),
         high_chol = as_factor(HighChol),
         heavy_alc = as_factor(HvyAlcoholConsump),
         smoker = as_factor(Smoker)
         healthcare = as_factor(AnyHealthcare),
         gen_health = as_factor(GenHlth),
         .keep="unused")

#Check for missing data
colSums(diabetes_ad,is.na)
```

## Exploratory Analysis

## Link to modeling page

[Click here for the modeling page.](Modeling.html "Modeling.html")
